---
title: "Markov_chains_and_processes"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# 1 - Implementation Markov Chains
    
### Steps 1 & 2 - Declaring a parameter space/initial state



```{r simulation_param_function}
generate_params <- function(n, t, print_probas=T){
  # Generates a list of jump probabilities from state 0 to 1, 
  # and vice-versa, plus all required parameters to simulate
  # subsequent Markov chains.
  params = list("init"=0,"n"=n,"t"=t,
                "p0to1"=round(runif(1,0,1),2), 
                "p1to0"=round(runif(1,0,1),2))
  if (print_probas) {
    cat("Probability jump 0 -> 1:", params$p0to1,"\n")
    cat("Probability jump 1 -> 0:", params$p1to0)
  }
  params
}
```


```{r example_params}
generate_params(n = 10, t = 20)
```


### Step 3 - Generating Markov chains



```{r generation_functions}
simulation_v1_iterative <- function(params) {
  # Initializes a matrix to the init value found in the variable params
  # rows = time steps, columns = unique markov chain
  chains = matrix(params$init, nrow=params$t, ncol=params$n)
  # Declares a function that performs a check condition (i.e. to 
  # determine whether to jump state or not)
  switch <- function(prob) {runif(1,0,1)<=prob}
  # Iterates over the matrix, updating when needed. Iterating over the 
  # rows starts at index 2 as we don't expect jumps at t=0 (1 in R) as it
  # is the fixed initial state.
  for (row in seq(2,params$t,1)) {
    for (col in seq(1,params$n,1)){
      # Updates the value dependent on the cell value at t-1 and
      # the result of the switch function
      if (chains[row-1,col]==0 && switch(params$p0to1)) {
        chains[row,col]=1
      } else if (chains[row-1,col]==1 && switch(params$p1to0)) {
        chains[row,col]=0
      } else {
        chains[row,col]=chains[row-1,col]
      }
    }
  }
  chains
}
simulation_v2_bernoulli_seqs <- function(params) {
  # Initializes a matrix to the init value found in the variable params
  # rows = time steps, columns = unique markov chain
  chains = matrix(params$init, nrow=params$t,   ncol=params$n)
  # Declares a function to more explicitly state that
  # a sequence of bernoulli values from 0 to 1 are 
  # generated
  generate_bernoulli_seq <- function(p,n) {rbinom(n, 1, p)}
  # Iterates over the matrix, updating when needed
  for (col in seq(1,params$n,1)) {
    # Generates two sequences Y and Z
    # we only need t-1 draws as we only update <chains>
    # starting at the index 2
    Y = generate_bernoulli_seq(params$p0to1, params$t-1)
    Z = generate_bernoulli_seq(params$p1to0, params$t-1)
    for (row in seq(1,params$t-1,1)) {
        chains[row+1, col] = chains[row,col]*(1-Z[row]) +
                             (1-chains[row,col])*Y[row]
      }
    }
  chains
}
```


```{r timing_generation_functions}
params = generate_params(n=1000, t=10000)
system.time(simulation_v1_iterative(params))
system.time(simulation_v2_bernoulli_seqs(params))
```

We see that the second function is faster by c. one order of magnitude. This can be explained by the absence of branching condition (i.e. `if` blocks) in the second function's body. Branching conditions are time-consuming operations for a CPU, requiring the use of [branch predictors](https://en.wikipedia.org/wiki/Branch_predictor) that can, and in this case do, waste time. As such, we will use the second function to generate Markov chains.

To generate Markov Chains, we arbitrarily set $t=25$ for data visualization purposes (we want to plot the data). Though we could set $n=4$ to get a single $4\times25$ Markov chain matrix, we would like to display different examples of generated Markov chains. As such, we simply run the simulation function four times with different parameter sets.

```{r generating_and_plotting_markov_chains, out.width="100%"}
par(mfrow=c(2,2))
for (i in 1:4){
  # Generates a new Markov chain with new parameters p0to1 and p1to0
  params = generate_params(n=1, t=25, print_probas=F)
  markov_chain = simulation_v2_bernoulli_seqs(params)
  # Plots the Markov chain
  title = paste("Generated Markov Chain n'", i, 
                "\np_0to1 =", params$p0to1, 
                "; p_1to0 =", params$p1to0)
  plot(markov_chain, type="l",
       main=title, xlab="Timesteps", ylab="Value (either 0 or 1)")
}
```


## 2 - Simulating a Markov Process


### Step 1 - Simulating $(Y_n)$



```{r markov_proc_params}
generate_MP_params <- function(n, t, a, b, d) {
  # Generates a list of probability of jumping from
  # state 0 to 1, and vice-versa
  # args:   n -> the number of chains to generate
  #         t -> the length of each chain
  params = list("init"=0,"n"=n,"t"=t,
                "alpha"=a,"beta"=b,"delta"=d,
                "p0to1"=a*d, "p1to0"=b*d)
  params
}
```



```{r simulating_Yn}
alpha = 5
beta = 10
delta = 0.01
n = 1000
t = 5000
MP_params = generate_MP_params(n, t, alpha, beta, delta)
Yn_simulations = simulation_v2_bernoulli_seqs(MP_params)
```



```{r visualizing_yn_simulations, out.width="100%"}
par(mfrow=c(2,2))
for (i in 1:4){
  plot(Yn_simulations[1:500,i], type="l",
       main=paste("First 500 steps of Markov Chain n'", i, 
                  "\nwith alpha=5, beta=10, delta=0.01"),
       xlab="timesteps", ylab="value (either 0 or 1)")
}
```



### Step 2 - Evaluating the distribution of $\tau_1^\delta=N_1\delta$



```{r retrieving_indexes}
fetch_indexes <- function(Yn, d) {
  # Retrieves the first index where 
  fetch_idx_one_simulation <- function(x){
    (c(which(x==1))[1]-1)*d # note the multiplication by d
    # we need to rescale by d as it is a frequency acquisition 
    # parameter. It comes handy for comparing distribution
    # with different d parameters afterwards
  }
  apply(Yn,2,fetch_idx_one_simulation)
}
```



```{r example_MP}
t1_simulation = fetch_indexes(Yn_simulations, delta)
```

We now have all the list of indexes of $N_1$ (of the first state jump from 0 to 1) of our previously declared Markov process approximations $(Y_n)$. We can then evaluate the distribution of $\tau^\delta_1$ where $\delta=0.01$.

```{r kd_plot_t1_simulations, out.width="100%"}
# Kernel density estimation
d <- density(t1_simulation)
# Kernel density plot
plot(d, lwd = 2, ylab="",
     main = paste("Distribution of Tau_1^delta\n", "Given delta =", delta),)
```


 
### Step 3 - Visualizing the empirical distribution of $\tau_1^\delta$



```{r delta_space_params}
n = 5000
t = 10000
alpha = 5
beta = 10
delta_sequence = as.matrix(c(1e-1,5e-2,1e-2,5e-3,1e-3,5e-4))
```

We want to iterate our simulation process over the sequence of $\delta$ parameters. To do so we declare the function `mapper`, which can be mapped over the sequence. `mapper` holds the composition of previously declared functions `generate_MP_params`, `simulation_v2_bernoulli_seqs`, and `fetch_indexes`.

As such, we generate the following simulations:

| parameters | Sim. 1 | Sim. 2 | Sim. 3 | Sim. 4 | Sim. 5 | Sim. 6 |
| --- | --- | --- | --- | --- | --- | --- | 
| n | 5000 | 5000 | 5000 | 5000 | 5000 | 5000 |
| t | 10000 | 10000 | 10000 | 10000 | 10000 | 10000 |
| $\alpha$ | 5 | 5 | 5 | 5 |  5 | 5 |
| $\beta$ | 10 | 10 | 10 | 10 | 10 | 10 |
| $delta$ | 0.1 | 0.05 | 0.01 | 0.005 | 0.001 | 0.0005 |

```{r delta_space_simulations}
mapper <- function(n, t, a, b, d) {
  # Function composition to be mapped over the sequence of delta
  # parameters s.t. fetch . simulate . parameter_generation()
  fetch_indexes(
    simulation_v2_bernoulli_seqs(
      generate_MP_params(n, t, a, b, d)), d)
}
t1_simulations = apply(delta_sequence, 1, function(x) {mapper(n,t,alpha,beta,x)})
```



```{r density_comparison_different_deltas, out.width="100%"}
# Declares the exponential distribution against which to plot 
# the simulated Tau_1 times
exponential_distribution = density(rexp(n, alpha))
# Declares the 2 by 3 plot area
par(mfrow=c(2,3), mai = c(0.4, 0.25, 0.3, 0.1))
for (i in 1:6){
  # Computes the kernel density for each simulation
  d <- density(t1_simulations[,i])
  # Plots said density against the exponential distribution
  plot(d, lwd = 1, col="blue",
       main = paste("Dist. of Tau_1 with delta=", delta_sequence[i], 
                    "\nvs. Exp(alpha=", alpha, ")", sep=""),
       xlab="", ylab="")
  lines(exponential_distribution, lw=2, lty=3, col="red")
  legend("topright", legend = c("Tau_1 dist.", 
                                paste("Exp(", alpha, ") dist.",sep="")),
       lwd = 3, col = c("blue", "red"))
}
```


