---
title: "kernel_estimator_selection_bootstrapping"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# 1 - Cross-Validation For Kernel Estimators

**A** -- Let's implement a function that computes, for a given vector $u$ and a given window $h$, the value $\hat{f}_h(.)$ for each coordinate of the vector $u$. Then let's simulate exponential variables and apply the kernel estimator with different values for $h$.

**B** -- Choosing $h$ can be done with cross-validation: Let's rely on the least-square contrast: $$C(g) = -\frac{2}{n}\sum^n_{i=1}g(X_i) + \int g(x)^2dx$$ Such that $\mathbb{E}[C(g)]$ is minimal when $f=g$ where $g$ is a candidate density. Let's demonstrate that state.

**C** -- Using the function `integrate`, let's compute the integral of the kernel estimator $f$ to the square and implement a function that computes the least square contrast on a different sample than the one previously simulated.

**D** -- Let's implement the hold-out estimator: The data is cut in half at random, half to be used for estimation, and the other half for the selection of the bandwidth parameter $h$. Then let's plot the corresponding estimator.

## A

Your code is reproduced here with comments below.

```{r}

f_h=rep(1,100)
h=0.15
x=rexp(100,2)
u=c(h,x)
for(sim in 1:100){
  n=1:100
  #K=(exp*(((-x)^2)*2)/(sqrt(2*pi))) <- declaration error (see comments), 
  #                                     exp() is a standard library function in
  #                                     R and not a variable.
  #f_h[sim]=(1/n*h)*sum(K*((u-X_i)/h)) <- X_i is not declared which breaks the
  #                                       code
}

#plot(density(f_h),bty='n') <- because the loop computation fails, it outputs 
#                              a faux-normal distribution

```

1. ``K=(exp*(((-x)^2)*2)/(sqrt(2*pi)))`` is erroneous because of the order of operations. Exponents have priority over multiplication so $K(x)=exp(-x^2/2)/\sqrt{2\pi}$ should be ``K = function(x) {exp(-(x^2)/2)/sqrt(2*pi)}``.

2. $u$, $\forall i\in\{1,...,n\}, X_i$ are unknown vectors that are input to a function (requested by the instruction, along with a real-valued bandwidth $h$). 

3. The idea behind the question is to create a convolution operation (the [kernel estimator](https://en.wikipedia.org/wiki/Kernel_density_estimation)) that performs an operation over slices of input signals (here the interspike intervals $X_i$) to approximate an underlying distribution law (here exponential):
    - The use of a bandwidth (like in signal processing, think radio signals) is meant to extract details from an input at different levels of details (think zooming in and out with a microscope). 
    - Since we only have one law underlying the data $X_i$, a single bandwidth $h$ should be necessary in the case of the exercise. This is why the latter questions wanted to implement differnt model selections, to find the best bandwidth $h$ to yield the actual exponential function that characterizes the ISI generated in this question.

## B

The demonstration is correct and exhaustive.

## C

The process described on paper looks about right, though the instruction requested it to be implemented in R. 

## D

It seems that you mentioned that the question D was also covered by your answer in B. I did not understand what you meant by writing ``1) b)d)``

# 2 - Parametric Bootstrap

## Overview 

Let's implement parametric bootstrap.

**A** -- Let's simulate $n = 50$ exponential variables with parameter $\theta_0 = 2$. This will be considered as observations.

## A

Completed as requested.