---
title: "EM algorithm for handling missing data"
author: "Aude Sportisse"
date: "25/10/2021"
output: html_document
---

```{r}
library(mvtnorm) #library for multivariate normal density
library(ggplot2) #library to have nice graphics
```


### Code an EM algorithm

We consider $X\sim \mathcal{N}(\mu,\Sigma)$, with
    $$\mu=\begin{pmatrix} 5 \\ -1
    \end{pmatrix} \textrm{ and } \Sigma=\begin{pmatrix} 1 & 0.5 \\ 0.5 & 1
    \end{pmatrix}.$$ We want to introduce $r=30\%$ of missing values in the variable $X_2$. We consider that the missing-data mechanism is MCAR.
  

**Q1)** Generate a bivariate normal set of sample size $n=100$, with mean $\mu$ and covariance matrix $\Sigma$ (use the package mvtnorm).

```{r}
n = 100
mu = c(5,-1)
Sigma = matrix(c(1, 0.5, 0.5, 1), ncol=2, nrow=2)
X = rmvnorm(n,mu,Sigma)
head(X)
```


**Q2)** Introduce MCAR missing values in $X_2$. 

```{r}
missing_idx.mcar <- sample.int(n,0.3*n) #indexes of values which will be missing
XNA <- X
XNA[missing_idx.mcar,2] <- NA
head(XNA)
```


The goal is now to estimate the parameters $\mu$ and $\Sigma$ in presence of missing values in $X_2$ by using the EM algorithm.

**Q3)** Propose a simple initialization for the EM algorithm. 

```{r}
#we have to estimate mu and Sigma
hat_mu <- apply(XNA,2,mean,na.rm=TRUE)
hat_mu
hat_Sigma <- cov(XNA,use="complete.obs") 
hat_Sigma <- var(XNA,na.rm=TRUE)
hat_Sigma
```


**Q4)** Write a function for the E-step and the M-step. 

```{r}
Estep=function(X, mu, Sigma, missing_idx)
{
n=nrow(X) 

#all the elements in X1 are observed
s1_vec = X[,1]
s11_vec = X[,1]^2

s2_vec = rep(0, n)
s22_vec = rep(0, n)

#for observed elements in X2
#setdiff(1:n, missing_idx): observed elements
s2_vec[setdiff(1:n, missing_idx)] = X[setdiff(1:n, missing_idx),2]
s22_vec[setdiff(1:n, missing_idx)] = X[setdiff(1:n, missing_idx),2]^2

#for missing elements in X2
s2_vec[missing_idx] = mu[2]+(Sigma[1,2]/Sigma[1,1])*(X[missing_idx,1]-mu[1])
s22_vec[missing_idx] = s2_vec[missing_idx]^2 + Sigma[2,2] - Sigma[1,2]^2/Sigma[1,1]


s12_vec = s1_vec*s2_vec

return(list(s1=sum(s1_vec), s2=sum(s2_vec), s11=sum(s11_vec), s22=sum(s22_vec), s12=sum(s12_vec)))
}

Mstep=function(X, s1, s2, s11, s22, s12)
{
n=nrow(X)
mu1=s1/n
mu2=s2/n
sigma1=s11/n-mu1^2
sigma2=s22/n-mu2^2
sigma12=s12/n-mu1*mu2
mu=c(mu1,mu2)
Sigma=matrix(c(sigma1, sigma12,sigma12,sigma2), nrow=2)
return(structure(list(mu=mu, Sigma=Sigma)))
}
```


**Q5)** Use the EM algorithm for $50$ iterations to estimate $\mu$ and $\Sigma$. Show the results. 

```{r}
for(i in 1:50)
{
# E step
E=Estep(XNA, hat_mu, hat_Sigma, missing_idx.mcar)
s1=E$s1
s11=E$s11
s2=E$s2
s22=E$s22
s12=E$s12
# M step
M=Mstep(XNA, s1, s2, s11, s22, s12)
hat_mu=M$mu
hat_Sigma=M$Sigma
}
```

```{r}
hat_mu
hat_Sigma
```


### Other questions

**Q6)** Vary $n$ and the percentage of missing values $r$.

**Q7)** We have estimated the parameters $\mu$ and $\Sigma$, can we impute the missing values? Try it!

**Q8)** Do you think the algorithm will still work for MNAR data? If you have the time, try it!

**Q9)** How to stop the EM algorithm? (other than by giving a predefined number of steps) 




