---
title: "EM algorithm for handling missing data"
author: "Aude Sportisse"
date: "25/10/2021"
output: html_document
---

# Code an EM algorithm

We consider $X\sim \mathcal{N}(\mu,\Sigma)$, with
    $$\mu=\begin{pmatrix} 5 \\ -1
    \end{pmatrix} \textrm{ and } \Sigma=\begin{pmatrix} 1 & 0.5 \\ 0.5 & 1
    \end{pmatrix}.$$ We want to introduce $r=30\%$ of missing values in the variable $X_2$. We consider that the missing-data mechanism is MCAR.
    
## Library imports

```{r}

library(mvtnorm)
library(pgmm)
library(mvtnorm)
library(mclust)
library(ggplot2)

```

## Distribution parameters
```{r}

mu = matrix(c(5, -1), nrow=1, ncol=2)
sigma = matrix(c(1, 0.5, 0.5, 1), nrow=2, ncol=2)

mu
sigma

```

## Questions

**Q1)** Generate a bivariate normal set of sample size $n=100$, with mean $\mu$ and covariance matrix $\Sigma$ (use the package mvtnorm).

```{r}

n = 100
data = rmvnorm(n, mu, sigma)

```

**Q2)** Introduce MCAR missing values in $X_2$. 

The goal is now to estimate the parameters $\mu$ and $\Sigma$ in presence of missing values in $X_2$ by using the EM algorithm.

```{r}

#p = 1/10
#binomial_draws = rbinom(n, 1, p)
#data[which(binomial_draws==1),2]=NA

missing_idx.mcar <- sample.int(n, 0.3*n)
X = data
X[missing_idx.mcar, 2] = NA

```

**Q3)** Propose a simple initialization for the EM algorithm. 

```{r}
n
# We use the empirical mean and sigma

mu_init = apply(X, 2, mean, na.rm=TRUE)
sigma_init = cov(X, use="complete.obs")
#sigma_init = var*X, na.rm=TRUE)

```

**Q4)** Write a function for the E-step and the M-step. 

```{r}

expectationStep <- function(X, mu, Sigma) {
  ######
  #### Expectation step of a bivariate dataset with missing data in the
  #### second dimension, i.e., X[,2]
  ######
    
  n=nrow(X)
  missing_idx = which(is.na(X[,2]))
  
  #all the element in X1 are observed
  s1_vec = X[,1]
  s11_vec = X[,1]^2
  s2_vec = rep(0,n)
  s22_vec = rep(0,n)
  
  # For observed elements in X2
  s2_vec[setdiff(1:n, missing_idx)] = X[setdiff(1:n, missing_idx), 2]
  s22_vec[setdiff(1:n, missing_idx)] = X[setdiff(1:n, missing_idx), 2]^2
  
  # For missing elements in X2
  s2_vec[missing_idx] = mu[2] + 
    (Sigma[1,2]/Sigma[1,1]) * (X[missing_idx,1]-mu[1])
  s22_vec[missing_idx] = s2_vec[missing_idx]^2 + 
    Sigma[2,2] - 
    Sigma[1,2]^2 / Sigma[1,1]
  
  s12_vec = s1_vec * s2_vec
  
  return(list(
    s1=sum(s1_vec), 
    s2=sum(s2_vec), 
    s11=sum(s11_vec), 
    s22=sum(s22_vec), 
    s12 = sum(s12_vec)
    )
  )
}

maximizationStep <- function() {
  
}



```

**Q5)** Use the EM algorithm for $50$ iterations to estimate $\mu$ and $\Sigma$. Show the results.

### Other questions

**Q6)** Vary $n$ and the percentage of missing values $r$ and see what happens. 

**Q7)** We have estimated the parameters $\mu$ and $\Sigma$, can we impute the missing values? Try it!

**Q8)** Do you think the algorithm will still work for MNAR data? If you have the time, try it!

**Q9)** How to stop the EM algorithm? (other than by giving a predefined number of steps) 




