---
title: "Network - Lesson 1"
author: "Pr. Charles Bouveyron"
date: "Last updated on `r Sys.Date()`"
output:
  html_document:
    toc: yes
    toc_float:
      collapsed: yes
      smooth_scroll: yes
    toc_depth: 2
    number_sections: yes
  pdf_document:
    toc: yes
    toc_depth: '2'
---

# Manipulation and characterization of networks


```{r}
library(igraph)

A = cbind(c(0,1,1,1),
          c(1,0,1,1),
          c(1,1,0,0),
          c(1,0,1,0))
```

By default, the graph.adjacency function sees the networks as directed.

```{r}
g = graph.adjacency(A)
plot(g)
degree(g)
```

To inform the function that the network is undirected, we have to use the `mode` option:

```{r}
g = graph.adjacency(A,mode = "undirected")
plot(g)
degree(g)
```

Another statistic we can compute is the numbers of cliques.

> Definition: A n-clique is a set of n nodes that are all connected together directly.

```{r}
cliques(g)
```

Here, we just have 4 1-cliques and 3 2-cliques

> Exercise: write a small code which computes a signature for the network by counting how many cliques of each type we have in this network.

```{r}
signature <- function(g,Nmax = 10){
  cl = cliques(g)
  s = rep(0,Nmax)
  for (i in 1:length(cl)){
    l = length(cl[[i]])
    s[l] = s[l] + 1
  }
  barplot(s)
  return(s)
}

signature(g)
```

## The case of the Sampson monk network

```{r}
#install.packages("VBLPCM")
library(VBLPCM)
data(sampson)

A <- as.matrix.network.adjacency(samplike)
A
```

Let's first try to visualize this data set:

```{r}
g = graph.adjacency(A)
plot(g)
```

> Exercice: use the code above to compute the clique signature for the Sampson data.

```{r}
signature(g)
```

We can see here a summary of the global and local density / complexity of the network. Having in particular one 6-clique is an important information in this small network: there is a group of 6 persons who are all friends, in a network of 18 persons!

